---
title: "isch-ctrl-rmats_eda"
output: html_document
date: "2024-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(maser)
library(rtracklayer)
```

#Test 1 Outputs

```{r}
#Loading Data, Test 1
summary.df=read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/summary.txt")
se.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/SE.MATS.JCEC.txt")
ri.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/RI.MATS.JCEC.txt")
a3ss.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/A3SS.MATS.JCEC.txt")
a5ss.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/A5SS.MATS.JCEC.txt")
mxe.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/t5-post-directory/MXE.MATS.JCEC.txt")
```


```{r}
#Total ASE Frequencies (Calculated From JCEC)
barplot1 = barplot(summary.df$TotalEventsJCEC, names.arg=summary.df$EventType, 
                   main="Total Alternative Splicing Events Identified",
                   xlab="Event Type", ylab="Frequency", col="lightgreen", 
                   ylim = c(0, max(summary.df$TotalEventsJCEC) * 1.2))
text(x = barplot1,y = summary.df$TotalEventsJCEC, 
     labels = summary.df$TotalEventsJCEC, pos=3, cex=0.9)

```



```{r}
#Significant ASE Frequencies (Calculated From JCEC)
barplot2 = barplot(summary.df$SignificantEventsJCEC, names.arg=summary.df$EventType, 
                   main="Significant Alternative Splicing Events Identified",
                   xlab="Event Type", ylab="Frequency", col="lightblue", 
                   ylim = c(0, max(summary.df$SignificantEventsJCEC) * 1.2))
text(x = barplot2,y = summary.df$SignificantEventsJCEC, 
     labels = summary.df$SignificantEventsJCEC, pos=3, cex=0.9)
```

```{r}
ase.df = bind_rows(mutate(se.df, EventType = "SE"),
  mutate(ri.df, EventType = "RI"), mutate(a3ss.df, EventType = "A3SS"),
  mutate(a5ss.df, EventType = "A5SS"), mutate(mxe.df, EventType = "MXE"))


#Filtering: 
  #read count is >= 10 for both sample groups, 
  #PSI is between 0.05 and 0.95
  #FDR <= 0.01
  #Delta PSI >= 0.05
  #P Value <= 0.05 
filter.ase = ase.df %>%
  filter((IJC_SAMPLE_1 + SJC_SAMPLE_1) / 2 >= 10, (IJC_SAMPLE_2 + SJC_SAMPLE_2) / 2 >= 10,
         IncLevel1 >= 0.05, IncLevel1 <= 0.95, IncLevel2 >= 0.05, IncLevel2 <= 0.95, 
         FDR <= 0.01, abs(IncLevelDifference) >= 0.05, PValue <= 0.05)

table(ase.df$EventType)
table(filter.ase$EventType)
```


```{r}
barplot3.df <- filter.ase %>%
  group_by(EventType) %>%
  summarise(Count = n())

ggplot(barplot3.df, aes(x = EventType, y = Count, fill = EventType)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  labs(title = "Alternative Splicing Events (Filtered)",
       x = "Event Type", y = "Count", fill = "Event Type") +
  theme_minimal() +
  theme(legend.position = "none")

barplot3.df
```



```{r}
#ASE Frequencies in Filtered Data: Higher Inclusion in Sample 1 vs Higher Inclusion in Sample 2
#Sample groups separated using IncLevelDifference (difference in average inclusion level between the two groups)
barplot4 <- filter.ase %>%
  mutate(SampleGroup = ifelse(IncLevelDifference > 0, "Sample 1", "Sample 2")) %>%
  group_by(EventType, SampleGroup) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(barplot4, aes(x = EventType, y = Count, fill = SampleGroup)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Stacked Bar Graph: Sample Group Bias by Event Type",
       x = "Event Type", y = "Count", fill = "Sample Group") +
  scale_fill_manual(values = c("Sample 1" = "red", "Sample 2" = "blue"),
                    labels = c("Ischemic (Sample Group 1)", "Control (Sample Group 2)")) +
  theme_minimal()

barplot4
```



```{r}
#Violin Plots with overlayed boxplot: Show the density and spread of delta psi values for each event type
ggplot(filter.ase, aes(x = EventType, y = IncLevelDifference, fill = EventType)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), alpha = 0.5, outlier.shape = NA) +
  labs(title = "ΔPSI Distribution for Significant Events",
       x = "Event Type", y = "ΔPSI (Sample 1 - Sample 2)") +
  theme_minimal() +
  theme(legend.position = "none")
```

A5SS and RI events were identified much more frequently in sample 1, ischemic, than sample 2, healthy. MXE and A3SS were also identified in ischemic samples more frequently than control samples. Skipped exon events were identified over a large range but more frequently in the control samples then ischemic samples.

```{r}
se.filtered <- se.filtered %>%
  mutate(SampleGroup = ifelse(IncLevelDifference > 0, "Ischemic", "Control"))


significant_se_genes <- se.filtered %>%
  select(GeneID, IncLevelDifference, FDR, SampleGroup)

gene_counts <- significant_se_genes %>%
  group_by(SampleGroup) %>%
  summarise(Count = n())

gene_counts



ggplot(se.filtered, aes(x = SampleGroup, y = IncLevelDifference, fill = SampleGroup)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), alpha = 0.5, outlier.shape = NA) +
  labs(title = "ΔPSI Distribution for Skipped Exon Events",
       x = "Sample Group", y = "ΔPSI (Sample 1 - Sample 2)") +
  theme_minimal() +
  scale_fill_manual(values = c("Ischemic" = "red", "Control" = "blue"))
```



#Test 3 Outputs
##Dataset Overview

```{r}
t3summary.df=read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/summary.txt")
```


```{r}
#Total ASE Frequencies (Calculated From JCEC)
t3barplot1 = barplot(t3summary.df$TotalEventsJCEC, names.arg=t3summary.df$EventType, 
                   main="Total Alternative Splicing Events Identified",
                   xlab="Event Type", ylab="Frequency", col="blue", 
                   ylim = c(0, max(t3summary.df$TotalEventsJCEC) * 1.2))
text(x = t3barplot1,y = t3summary.df$TotalEventsJCEC, 
     labels = t3summary.df$TotalEventsJCEC, pos=3, cex=0.9)


#Significant ASE Frequencies (Calculated From JCEC)
t3barplot2 = barplot(t3summary.df$SignificantEventsJCEC, names.arg=t3summary.df$EventType, 
                   main="Significant Alternative Splicing Events Identified",
                   xlab="Event Type", ylab="Frequency", col="lightblue", 
                   ylim = c(0, max(t3summary.df$SignificantEventsJCEC) * 1.2))
text(x = t3barplot2,y = t3summary.df$SignificantEventsJCEC, 
     labels = t3summary.df$SignificantEventsJCEC, pos=3, cex=0.9)
```


##Skipped Exons
```{r}
t3se.df = read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/SE.MATS.JCEC.txt")
```

```{r}
#Averaging Read Counts
average_from_comma <- function(x) {
  sapply(strsplit(as.character(x), ","), function(vals) mean(as.numeric(vals)))
}

#Filtering Data
  #Read count >= 10 for both sample groups
  #PSI is between 0.05 and 0.95
  #FDR <= 0.01 
  #Delta PSI >= 0.05
t3se.df.filtered <- t3se.df %>%
  mutate(IJC_SAMPLE_1_avg = average_from_comma(IJC_SAMPLE_1),
         SJC_SAMPLE_1_avg = average_from_comma(SJC_SAMPLE_1),
         IJC_SAMPLE_2_avg = average_from_comma(IJC_SAMPLE_2),
         SJC_SAMPLE_2_avg = average_from_comma(SJC_SAMPLE_2)) %>%
  filter(IJC_SAMPLE_1_avg >= 10, SJC_SAMPLE_1_avg >= 10,
         IJC_SAMPLE_2_avg >= 10, SJC_SAMPLE_2_avg >= 10,
         IncLevel1 >= 0.05, IncLevel1 <= 0.95,
         IncLevel2 >= 0.05, IncLevel2 <= 0.95,
         FDR <= 0.01, abs(IncLevelDifference) >= 0.05, PValue <= 0.05)
```


```{r}
#Delta PSI Distribution in Significant Skipped Exon Events
hist(t3se.df.filtered$IncLevelDifference, 
     main="Distribution of ΔΨ for Significant SE Events", 
     xlab="ΔΨ (Inclusion Level Difference)", 
     col="lightgreen", breaks=20)
```




## Maser

```{r}
library(maser)
library(rtracklayer)
```

```{r}
#loading rmats outputs into maser
path <- "/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory"
data <- maser(path, c("Ischemic", "Control"), ftype = "JCEC")

summary_data <- summary(data, type = "SE")
head(summary_data)

filtered_data <- filterByCoverage(data, avg_reads = 10)
significant_data <- topEvents(filtered_data, fdr = 0.01, deltaPSI = 0.05)
```


```{r}
#Visualizing significant SE Events
volcano(filtered_data, fdr = 0.01, deltaPSI = 0.05, type = "SE")

dotplot(significant_data, type = "SE")

pca(filtered_data, type = "SE")

boxplot_PSI_levels(filtered_data, type = "SE")
```



## BED Files

```{r}
#RMATS to BED SE
rMATStoBEDSE <- function(rmats_line) {
  event <- strsplit(rmats_line, "\t")[[1]]
  
  ID <- event[1]
  chr <- event[4]
  strand <- event[5]
  exonStart <- as.integer(event[6])
  exonEnd <- as.integer(event[7])
  upstreamES <- as.integer(event[8])
  upstreamEE <- as.integer(event[9])
  downstreamES <- as.integer(event[10])
  downstreamEE <- as.integer(event[11])
  
  #block sizes
  UEsize <- upstreamEE - upstreamES
  DEsize <- downstreamEE - downstreamES
  MEsize <- exonEnd - exonStart
  
  #block starts
  UEstart <- 0
  MEstart <- exonStart - upstreamES
  DEstart <- downstreamES - upstreamES
  
  #define BED files
  chrom <- chr
  chromStart <- upstreamES
  chromEnd <- downstreamEE
  name <- ID
  score <- 500
  thickStart <- upstreamES
  thickEnd <- downstreamEE
  itemRgb <- "255,0,0"
  
  #inclusion bed12 line
   incBlockCount <- 3
  incBlockSizes <- paste(UEsize, MEsize, DEsize, sep = ",")
  incBlockStarts <- paste(UEstart, MEstart, DEstart, sep = ",")
  inclusion_track <- paste(
    chrom, chromStart, chromEnd, name, score, strand, thickStart, thickEnd,
    itemRgb, incBlockCount, incBlockSizes, incBlockStarts, sep = "\t"
  )

  #exclusion bed12 line
  excBlockCount <- 2
  excBlockSizes <- paste(UEsize, DEsize, sep = ",")
  excBlockStarts <- paste(UEstart, DEstart, sep = ",")
  exclusion_track <- paste(
    chrom, chromStart, chromEnd, name, score, strand, thickStart, thickEnd,
    itemRgb, excBlockCount, excBlockSizes, excBlockStarts, sep = "\t"
  )
  
  
   cat("track name=\"Inclusion Track\"\n", inclusion_track, "\n", sep = "")
  cat("track name=\"Exclusion Track\"\n", exclusion_track, "\n", sep = "")
}
```


```{r}
test_row = t3se.df.filtered[1, ]
test_line <- paste(test_row, collapse = "\t")

rMATStoBEDSE(test_line)
```

```{r}
event_1 = t3se.df.filtered %>%
  slice_max(order_by = abs(IncLevelDifference), n = 1)

event_line_1 = paste(event_1, collapse = "\t")

sink("most_significant_event.bed")
rMATStoBEDSE(event_line_1)
sink()
```





```{r}
se.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/SE.MATS.JCEC.txt")
ri.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/RI.MATS.JCEC.txt")
a3ss.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/A3SS.MATS.JCEC.txt")
a5ss.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory/A5SS.MATS.JCEC.txt")
mxe.df <- read_tsv("/project/cphg-millerlab/katherine_shillito/isch-ctrl-rmats/04_post-outputs/test3-post-directory//MXE.MATS.JCEC.txt")

all.events.df = bind_rows(mutate(se.df, EventType = "SE"),
  mutate(ri.df, EventType = "RI"), mutate(a3ss.df, EventType = "A3SS"),
  mutate(a5ss.df, EventType = "A5SS"), mutate(mxe.df, EventType = "MXE"))

all.events.df <- all.events.df %>%
  separate(IncLevel1, into = c("IncLevel1_1", "IncLevel1_2", "IncLevel1_3"), sep = ",", convert = TRUE) %>%
  separate(IncLevel2, into = c("IncLevel2_1", "IncLevel2_2", "IncLevel2_3"), sep = ",", convert = TRUE)

#all.events.df <- all.events.df %>%
 # mutate(
 #   IncLevel1 = as.numeric(IncLevel1),
  #  IncLevel2 = as.numeric(IncLevel2)
 # )

#all.events.df <- all.events.df %>%
  #mutate(AverageReadCount = (IncLevel1 + IncLevel2) / 2)


#all.events.df <- all.events.df %>%
  #mutate(AveragePSI1 = rowMeans(select(., starts_with("IncLevel1")), na.rm = TRUE),
  #       AveragePSI2 = rowMeans(select(., starts_with("IncLevel2")), na.rm = TRUE))
```

```{r}
head(all.events.df)
str(all.events.df)
summary(all.events.df)
```


```{r}
# Check for NA values in the dataset
colSums(is.na(all.events.df))
```

```{r}
# Percentage of missing data per column
sapply(all.events.df, function(x) mean(is.na(x)) * 100)
```

```{r}
table(all.events.df$EventType)
```


```{r}
#Inspecting recommended filtering criteria
#average read count >= 10 in both sample groups
summary(all.events.df$AverageReadCount)
hist(all.events.df$AverageReadCount, main = "Average Read Count Distribution", xlab = "Average Read Count", breaks = 20)
#average PSI <0.05 or >0.95 in both sample groups
summary(all.events.df$AveragePSI)
hist(all.events.df$AveragePSI, main = "Average PSI Distribution", xlab = "Average PSI", breaks = 20)
#FDR <= 0.01
summary(all.events.df$FDR)
hist(all.events.df$FDR, main = "FDR Distribution", xlab = "FDR", breaks = 20)
#abs(deltaPSI) >= 0.05
summary(abs(all.events.df$IncLevelDifference))
hist(abs(all.events.df$IncLevelDifference), main = "ΔPSI Distribution", xlab = "Absolute ΔPSI", breaks = 20)


```






```{r}
library(maser)
library(rtracklayer)
```

```{r}

```


